---
title: 'Reproducible Research - CP #1'
author: "Marc-Antoine E."
date: "15 avril 2016"
output: html_document
---

R markdown document on first course project

Following analysis are based on data that can be downloaded at the following link:
https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2Factivity.zip

# Loading and preprocessing the data

Install requested packages if not found
```{r, echo = TRUE}
list.of.packages <- c("ggplot2", "dplyr", "lattice")
new.packages <- list.of.packages[!(list.of.packages %in% installed.packages()[,"Package"])]
if(length(new.packages)) install.packages(new.packages)
library(ggplot2)
library(dplyr)
library(lattice)
```

Unzip, read the data and store it in a new variable
Overview of the new variable
```{r, echo = TRUE}
unzip("repdata-data-activity.zip")
activity <- read.csv(file = "activity.csv")
head(activity)
```

# What is mean total number of steps taken per day?

Group by date
```{r, echo = TRUE}
activity_groupbyday <- data.frame(activity$date,activity$steps)
activity_groupbyday <- aggregate(activity_groupbyday$activity.steps, by=list(Date = activity_groupbyday$activity.date),FUN = sum )
names(activity_groupbyday) <- c("Date","Steps")
activity_groupbyday$ID <- as.numeric(rownames(activity_groupbyday))
activity_groupbyday <- activity_groupbyday[!is.na(activity_groupbyday$Steps),]
```

Print activity histogram
```{r, echo = TRUE}
ggplot(activity_groupbyday) + geom_histogram(aes(x=ID, weight = Steps), fill= "red", binwidth = 0.5) + ggtitle("Number of steps per day") + ylab("Steps [#]") + xlab("Day")
```

Calculate mean and median
```{r, echo = TRUE}
stat_activity <- c(mean(activity_groupbyday$Steps),median(activity_groupbyday$Steps))
names(stat_activity) <- c("Mean","Median")
stat_activity
```

# What is the average daily activity pattern?

Average daily activity pattern
```{r, echo = TRUE}
daily_pattern <- tapply(activity$steps,activity$interval,FUN = mean, na.rm = TRUE)
head(daily_pattern)
```
Print daily activity pattern
```{r, echo = TRUE}
plot(row.names(daily_pattern), daily_pattern, type = "l", xlab = "Time interval", ylab = "Av. number of steps", main = "Daily activity pattern", col = "blue")
```
5-minute interval, on average across all the days in the dataset, containing the maximum number of steps
```{r, echo = TRUE}
interval_max <- which.max(daily_pattern)
names(interval_max)
```

# Imputing missing values

Total number of missing values in the dataset (i.e. the total number of rows with NAs)
```{r, echo = TRUE}
number_NA <- sum(is.na(activity))
number_NA
```

Filling in all of the missing values in the dataset : mean for that 5-minute interval
```{r, echo = TRUE}
steps_new <- NULL
for (i in 1:nrow(activity)){
  if (is.na(activity[i,]$steps)){
    steps_new[i] <- daily_pattern[row.names(daily_pattern) == activity[i,]$interval]
  }
  else {
    steps_new[i] <- activity[i,]$steps
    }
}
head(steps_new)
```

Create a new dataset that is equal to the original dataset but with the missing data filled in
```{r, echo = TRUE}
activity_new <- activity
activity_new$steps <- steps_new
head(activity_new)
```
Group by date
```{r, echo = TRUE}
activity_groupbyday2 <- data.frame(activity_new$date,activity_new$steps)
head(activity_groupbyday2)
activity_groupbyday2 <- aggregate(activity_groupbyday2$activity_new.steps, by=list(Date = activity_groupbyday2$activity_new.date),FUN = sum )
names(activity_groupbyday2) <- c("Date","Steps")
activity_groupbyday2$ID <- as.numeric(rownames(activity_groupbyday2))
```
Print activity histogram
```{r, echo = TRUE}
ggplot(activity_groupbyday2) + geom_histogram(aes(x=ID, weight = Steps), fill= "red", binwidth = 0.5) + ggtitle("Number of steps per day #2") + ylab("Steps [#]") + xlab("Day")
```

Calculate mean and median
```{r, echo = TRUE}
stat_activity2 <- c(mean(activity_groupbyday2$Steps),median(activity_groupbyday2$Steps))
names(stat_activity2) <- c("Mean","Median")
stat_activity2
```

# Are there differences in activity patterns between weekdays and weekends?

Creating a new factor variable in the dataset with two levels - "weekday" and "weekend" indicating whether a given date is a weekday or weekend day
```{r, echo = TRUE}
for (i in 1:nrow(activity_new)){
  if ((weekdays(as.Date(activity_new$date[i])) == "dimanche"||weekdays(as.Date(activity_new$date[i])) == "samedi")){
    activity_new$week[i] <- "weekend"
  }
  else {
    activity_new$week[i] <- "weekday"
  }
}
head(activity_new)
```

Group by weekday and interval
```{r, echo = TRUE}
extract_activity <- aggregate(activity_new$steps, by =list(activity_new$interval, activity_new$week), data = activity_new,FUN = mean)
names(extract_activity) <- c("Interval","Week", "Steps")
```
Make a panel plot
```{r, echo = TRUE}
xyplot(Steps ~ Interval | Week, extract_activity, type = "l", layout = c(1,2), xlab = "Interval", ylab = "Steps")
```

